---
title: "MSB1015_Assignment 2"
author: "Suus ten Hage"
date: "25 september 2019"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---
**Project Description**
In this project I will train a Partial Least Square (PLS) algorithm to predict the bioling point of alkines from their structure. This is based on data from Wiener, 1947. 

** 0. Load packages**
If you haven't installed the packages on your computer yet, please uncomment the 
install.packages lines. 

```{r}
# install.packages("WikidataQueryServiceR")
library("WikidataQueryServiceR")
```

** 1. Get the data from WikiData.** 
In this section I use the WikidataQueryServiceR package to obtain all alkines with known boiling points from WikiData. To do this, I first select all chemical compounds (Q11173), their labels and their boiling points (P2102). Next, I select all compounds with a label that is in English. Finally, I select the compounds whose label ends with "ane" to select the alkanes. This results in 91 alkanes and their boiling points. 

```{r}
data = query_wikidata('SELECT DISTINCT ?alkanes ?alkaneLabel ?boilingPoint      WHERE {
    ?alkanes (wdt:P31|wdt:P279) wd:Q11173;
      rdfs:label ?alkaneLabel;
      wdt:P2102 ?boilingPoint.
    FILTER((LANG(?alkaneLabel)) = "en")
    FILTER(STRENDS(?alkaneLabel, "ane"))
  }'
)
```



