---
title: "MSB1015_Assignment 2"
author: "Suus ten Hage"
date: "25 september 2019"
output: html_notebook
---

**Project Description**
In this project I will train a Partial Least Square (PLS) algorithm to predict the bioling point of alkines from their structure. 

**0. Load packages**
If you haven't installed the packages on your computer yet, please uncomment the 
install.packages lines. 

```{r}
# install.packages("WikidataQueryServiceR")
library("WikidataQueryServiceR")
#install.packages('rJava')
library('rJava')
#install.packages('rcdk')
library('rcdk')
#install.packages('pls')
library('pls')

```

**1. Get the data from WikiData.** 
In this section I use the WikidataQueryServiceR package to obtain all simple alkines with known boiling points from WikiData.

```{r}
data = query_wikidata('SELECT ?comp ?compLabel ?bp ?bpUnit ?bpUnitLabel ?smiles WHERE {   
    ?comp wdt:P31/wdt:P279* wd:Q41581 ;
    p:P2102 [ps:P2102 ?bp ;           
            psv:P2102/wikibase:quantityUnit  
            ?bpUnit         
            ] ;
    wdt:P233 ?smiles.
    SERVICE wikibase:label { bd:serviceParam wikibase:language "[AUTO_LANGUAGE],en". } 
    } '
)

```

**3.Quantify the alkane structures**

```{r}

#Get descriptors
parsed_smiles <- parse.smiles(data$smiles, kekulise=TRUE)

descNames <- c(
'org.openscience.cdk.qsar.descriptors.molecular.APolDescriptor',
'org.openscience.cdk.qsar.descriptors.molecular.WienerNumbersDescriptor'
)

descriptors <- eval.desc(parsed_smiles, descNames)
```

**4. Data pre-processing**
Not all boiling points in the database are in kelvin, which is the SI unit for temperature. Before I continue, I make sure all temperatures are in kelvin. 

```{r}
#Find all rownumbers of the values in Celcius. 
celcius <-  which(grepl('degree Celsius', data$bpUnitLabel))
for (i in 1:length(celcius)) {
  #Convert the value to kelvin
  data$bp[celcius[i]] <- data$bp[celcius[i]] + 273.15
  #Convert the label to kelvin. 
  data$bpUnitLabel[celcius[i]] <- 'kelvin'
}

#Find all rownumbers of the values in Fahrenheit. 
fahrenheit <- which(grepl('degree Fahrenheit', data$bpUnitLabel))
for (i in 1:length(fahrenheit)) {
  #Convert the value to kelvin
  data$bp[fahrenheit[i]] <- (data$bp[fahrenheit[i]] + 459.67) * (5/9)
  #Convert the label to kelvin. 
  data$bpUnitLabel[fahrenheit[i]] <- 'kelvin'
}

```
